A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2014-11-22 journal.tex
Final draft

%Further, by applying simple extensions to the backoff mechanism of CSMA/ECA it is able to support many contenders in a collision-free schedule, surpassing the achieved throughput of CSMA/CA and providing short-term throughput fairness among contenders.

This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes. Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed

%This work describes CSMA/ECA and its extensions that allow it to achieve a collision-free schedule with many contenders in non-saturated conditions, as well as its performance when coexisting with CSMA/CA nodes. Further, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.

@@

%handle many contenders without sacrificing short-term throughput fairness.

@@

%where $k$ is reset ($k=0$) after each successful transmission and $m$ is the maximum backoff stage of typical value $m=5$

@@

	%greater throughput than CSMA/CA. Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA with nodes sending $2^m$ packets in each attempt. This behavior will be called CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+FS}}$ (Max Ag.) for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$, respectively from here forth.

@@

	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k} = 2^{k}$ and $l_{k-1} = 2^{k-1}$ packets respectively. These are given by:

	\begin{subequations}
		\begin{align}
			s_{k}(l_{k}) &= \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
			s_{k-1}(l_{k-1}) &= \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
		\end{align}
	\end{subequations}

@@

	The \emph{Lower-bound} is derived from considering the operation of an ideal CSMA/ECA$_{\text{Hys+FS}}$ network. Nodes use the minimum backoff stage possible and aggregate proportionally, thus yielding the minimum throughput achievable by a CSMA/ECA$_{\text{Hys+FS}}$ network. It is computed following~(\ref{eq:bound}).

	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and consequently the level of aggregation. It is also computed using~(\ref{eq:bound}) but considering $k=m$ for all nodes.

	The maximum throughput achievable is the result of deactivating the Fair Share rules by forcing nodes to use maximum aggregation regardless of their backoff stage. This is called \emph{Maximum Aggregation (Hys+MaxAg)} in Figure~\ref{fig:ECA-bounds-comparison}. It can be derived from~(\ref{eq:bound}) considering $l_{m} = 2^{m}$.

	%The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+MaxAg}}$. In opposition to the Lower-bound, here the nodes are assumed to aggregate as much as possible in spite of using the minimum schedule, hence yielding the maximum throughput the system can achieve. The Upper-bound is also computed using~(\ref{eq:bound}) but considering $l_{\alpha} = 2^{m}$, regardless of the backoff stage.
%		
	It is interesting to see in Figure~\ref{fig:ECA-bounds-comparison} how collisions force colliding CSMA/ECA$_{\text{Hys+FS}}$ contenders to increase their backoff stage and aggregate more with Fair Share. This explains why the CSMA/ECA$_{\text{Hys+FS}}$ curve separates itself from the \emph{Lower-bound} at a very low number of contenders.

@@

	\subsubsection{Time between successful transmissions}\label{timeBetweenSxTx}
	It is related to the elapsed time between the contention for transmission and the reception of an ACK.
	%the arrival of a packet to the MAC queue up until an acknowledgment is received for such packet(s).

@@

	In Figure~\ref{fig:serviceTime-sat} all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions. This is due to the multiple packets that are sent in each attempt. CSMA/CA$_{\text{MaxAg}}$, though, has an increased value due to collisions also taking longer channel time. %Furthermore, the variation seen in CSMA/ECA$_{\text{Hys+MaxAg}}$ is due to nodes having different backoff stages.

	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it achieves greater throughput than CSMA/CA (see Figure~\ref{fig:throughput-sat}).

	%CSMA/ECA$_{\text{Hys+FS}}$ shows a lower delay due to the aggregation performed via Hysteresis and Fair Share, mainly because these aggregated transmissions are acknowledged by a single ACK. It is important to highlight that because nodes are in saturation, the arrival time to the MAC queue is considered the same for all the aggregated packets.

@@

	%Further, although there is still a probability of collisions due to the presence of CSMA/CA contenders, these collisions trigger the Hysteresis and Fair Share extension of CSMA/ECA$_{\text{Hys+FS}}$ nodes, producing the throughput increase when compared to a CSMA/CA network. Figure~\ref{fig:mixedThroughput-sat-stations} provides an overview of the average throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations respectively.

%	\begin{figure*}[tb]
%		\centering
%		\includegraphics[width=0.4\linewidth,angle=-90]{figures/saturated/mixed/throughput-saturated-mixed-stationGroups.eps}
%		\caption{Average throughput of stations in a mixed network: a) 25\% of CSMA/CA nodes, b) 50\% of CSMA/CA nodes and c) 75\% of CSMA/CA nodes. The remaining proportion is composed of CSMA/ECA$_{\text{Hys+FS}}$ nodes}
%		\label{fig:mixedThroughput-sat-stations}
%	\end{figure*}

@@

The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput. Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput, the latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth. Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.

@@

\begin{itemize}
	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them. As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence would disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's. A similar effect is expected when in the presence of hidden nodes.
	%this problem is similar to that of hidden nodes. If the channel is found busy the nodes will freeze their respective backoff counters, while those unable to detect the channel busy with the overlapping AP transmission (or hidden node's) will not. This kind of behaviour results in slot drifts that in consequence would disrupt the collision-free schedule and approximate CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's.
	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
	\begin{itemize}
		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets. APs in CSMA/ECA$_{\text{Hys+FS}}$ can use a big TXOP in order to transmit more than users.
		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories. In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA. Priorities to the multiple queues can be granted through different minimum and maximum contention windows.
	\end{itemize}
	%Nevertheless, the AIFS mechanism can still be used. If a fixed AIFS is assigned for CSMA/ECA$_{\text{Hys+FS}}$ stations, the multiple queues and TXOP mechanisms will take care of prioritisation. Furthermore, using the shortest AIFS value would promote CSMA/ECA$_{\text{Hys+FS}}$ adoption by prioritising its traffic over EDCA's.
	\item Coexistence with EDCA stations in the same WLAN.
\end{itemize}

%\begin{itemize}
%	\item Multiple-User Multiple-Input Multiple-Output (MU-MIMO): nodes, like the AP for instance, would be able to send packets to different users using the same transmission slot; further increasing the aggregated throughput of the network.
%	\item Channel bonding: increasing the channel bandwidth also provides a throughput improvement, even though no modification to CSMA/ECA$_{\text{Hys+FS}}$ is required.
%\end{itemize}
%
%Apart from the mentioned above, there are other features that are not yet straightforward supported by CSMA/ECA$_{\text{Hys+FS}}$:
%
%\begin{itemize}
%	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS), would not work~\cite{jaumeTD}.
%	\begin{itemize}
%		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets. APs in CSMA/ECA$_{\text{Hys+FS}}$ can make use of the TXOP feature in order to transmit more than users.
%		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories. In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA.
%	\end{itemize}
%\end{itemize}
%
%Other WLANs features, like security, are MAC agnostic, so its implementation using CSMA/ECA$_{\text{Hys+FS}}$ is straightforward.
%
%The aforementioned features would level CSMA/ECA$_{\text{Hys+FS}}$ features with the CSMA/CA implementation for the current WLAN standard, providing sufficient insight to be considered a MAC replacement in WLANs.

>>>>

This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes. Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.

@@

@@

@@

@@

	where $s_{k}(l_{k})$ and $s_{k-1}(l_{k-1})$ are the throughput achieved by the nodes at the $k$-th and $(k-1)$-th backoff stages sending $l_{k}$ and $l_{k-1}$ packets respectively. These are given by:

	\begin{subequations}
		\begin{align}
			&s_{k}(l_{k}) = \frac{l_{k} L}{hT(l_{k})+2(N-h)T(l_{k-1}) + \sigma_{e}(C - N)}\label{eq:HighNodes}\\
			&s_{k-1}(l_{k-1}) = \frac{l_{k-1} L}{(N-h)T(l_{k-1}) + k\frac{T(l_{k})}{2}}\label{eq:LowNodes}
		\end{align}
	\end{subequations}

@@

	The \emph{Lower-bound} is derived from considering the operation of an ideal CSMA/ECA$_{\text{Hys+FS}}$ network. Nodes use the minimum backoff stage possible and aggregate proportionally, thus yielding the minimum throughput achievable by a CSMA/ECA$_{\text{Hys+FS}}$ network. It is computed following~(\ref{eq:bound}) with $l_{k}=2^{k}$ and $l_{k-1}=2^{k-1}$.

	The \emph{Upper-bound} is obtained from considering the operation of a network using CSMA/ECA$_{\text{Hys+FS}}$, but forcing nodes to use the maximum backoff stage for determining the cycle length and the level of aggregation. It is also computed using~(\ref{eq:bound}) but considering that all nodes are in the maximum backoff stage ($k=m$) and therefore $l_{k}=2^{m}$.

	The maximum throughput achievable is the result of deactivating the Fair Share rules by forcing nodes to use maximum aggregation regardless of their backoff stage. This is called \emph{Maximum Aggregation (Hys+MaxAg)} in Figure~\ref{fig:ECA-bounds-comparison}. It can be derived from~(\ref{eq:bound}) considering $l_{k} = 2^{m}$ and $l_{k-1}=2^{m}$.

	It is interesting to see in Figure~\ref{fig:ECA-bounds-comparison} how collisions force colliding CSMA/ECA$_{\text{Hys+FS}}$ contenders to increase their backoff stage and aggregate more with Fair Share. This explains why the CSMA/ECA$_{\text{Hys+FS}}$ curve separates itself from the \emph{Lower-bound} at a very low number of contenders.

@@

	\subsubsection{Time between successful transmissions}\label{timeBetweenSxTx}
	It is related to the elapsed time between the contention for transmission and the reception of an ACK.

@@

	In Figure~\ref{fig:serviceTime-sat} all tests with maximum aggregation, namely CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$, have an increased average time between successful transmissions. This is due to the multiple packets that are sent in each attempt. CSMA/CA$_{\text{MaxAg}}$, though, has an increased value due to collisions also taking longer channel time.

	Although CSMA/ECA$_{\text{Hys+FS}}$ has an increased average time between successful transmissions due to Fair Share, it has a lower metric when compared with the maximum aggregation curves in Figure~\ref{fig:serviceTime-sat}.

@@

@@

The current PHY enhancements considered by the HEW study group include higher modulation and coding schemes as well as full duplex radios, capable of receiving and transmitting at the same time, thus augmenting the achievable throughput. Furthermore, already existing technologies like Channel bonding and Multiple-User Multiple-Input Multiple-Output (MU-MIMO) also seek to increase the throughput. The latter by allowing the transmission of different packets to multiple destinations at the same time, while the former bonds multiple WiFi channels together in order to increase the available bandwidth. Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.

@@

\begin{itemize}
	\item The performance of a CSMA/ECA$_{\text{Hys+FS}}$ network in presence of other neighbouring wireless networks and hidden nodes: transmissions from other network may negatively affect the maintenance of the collision-free schedule when not all devices in the CSMA/ECA$_{\text{Hys+FS}}$ WLAN are able to listen to them. As a consequence, not all stations will pause the backoff accordingly, resulting in large slot drifts that in consequence will disrupt the collision-free schedule, approximating CSMA/ECA$_{\text{Hys+FS}}$ performance to CSMA/CA's. A similar effect is expected when in the presence of hidden nodes.
	\item Traffic differentiation: although priorities using different contention windows in CSMA/ECA$_{\text{Hys+FS}}$ proved to outperform CSMA/CA, other Enhanced Distributed Channel Access (EDCA) mechanisms like the Arbitration Inter-Frame Spacing (AIFS) would not work~\cite{jaumeTD}, whereas the following EDCA mechanisms would:
	\begin{itemize}
		\item Transmission Opportunity (TXOP): stations with an increased TXOP are able to transmit more packets. APs in CSMA/ECA$_{\text{Hys+FS}}$ can use a big TXOP in order to transmit more than users.
		\item Multiple Queues:~\cite{jaumeTD} provides performance metrics for two traffic categories. In order to transition to CSMA/ECA$_{\text{Hys+FS}}$ a total of four access categories (AC) should be implemented, as in EDCA. Priorities to the multiple queues can be granted through different minimum and maximum contention windows.
	\end{itemize}
	CSMA/ECA$_{\text{Hys+FS}}$ with multiple access categories is expected to provide better traffic differentiation in WLANs, mainly due to the elimination of collisions by using a deterministic backoff after each access category's successful transmission.
	\item Coexistence with EDCA stations in the same WLAN.
\end{itemize}
